---
title: "Oncoprint"
output: html_document
date: "2026-06-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Driver mutation heatmap 

pipeline? for data:

TCGA MC3 somatic mutations (raw MAF file)
          ↓
ANNOVAR  — adds functional annotations, SIFT, COSMIC, gnomAD etc.
          ↓
ABSOLUTE or PyClone — adds CCF, clonality estimates
          ↓
Custom driver calling script — adds is_driver_mut, pan_cancer_driver,
                               oncogene, tumor_suppressor columns
                               likely using Bailey et al. or COSMIC
                               as the reference driver list
***Overview***:

**Driver related**
pan_cancer_driver    → binary, is this mutation a pan-cancer driver?
is_driver_mut        → binary, is this a driver mutation?
driver_category      → category of driver
newDriverCat         → updated driver category
oncogene             → is the gene an oncogene?
tumor_suppressor     → is it a tumour suppressor?
Role                 → likely oncogene/TSG/both

**Mutation identity**
gene.refgene         → the gene name — this is what links to your expression data
cancertype           → cancer type per mutation
sampleid             → links to your expression data

**Mutation effect**
ccf                  → cancer cell fraction — is the mutation clonal or subclonal?
prob.clonal          → probability of being clonal
prob.subclonal       → probability of being subclonal
is_deleterious       → is the mutation damaging?


```{r}

possible_roots <- c("~/GenomeDK", "faststorage/Project", "")

smart_read <- function(relative_path, read_func = readRDS, ...) {
  full_paths <- file.path(possible_roots, relative_path)
  existing_path <- full_paths[file.exists(full_paths)][1]
  
  if (is.na(existing_path)) {
    stop(paste("No root for:", relative_path))
  }
  message("Loading: ", existing_path)
  return(read_func(existing_path, ...))
}

library(tidyverse)
library(broom)

drivers <- smart_read("thesis/data/TCGA_Derived_data/TCGA_annotated_driver_mutations.rds")
clin <- smart_read("thesis/data/TCGA_clinical/clinical_dataset.rds")

clin <- clin |> rename(sampleid = `_PATIENT`)

clin$BMI <- clin$weight / (clin$height/100)^2

clin <- clin |>
  mutate(bmi_groups = case_when(
    BMI < 18.5              ~ "Normal",
    BMI >= 18.5 & BMI < 25  ~ "Normal",
    BMI >= 25 & BMI < 30    ~ "Overweight",
    BMI >= 30               ~ "Obese",
    TRUE                    ~ NA_character_
  ), bmi_groups = factor(bmi_groups, levels = c("Normal", "Overweight", "Obese"))
  ) |> 
  mutate(age_cat = if_else(age_at_initial_pathologic_diagnosis < 65, "young","old" ),
         age_cat = factor(age_cat, levels = c("young", "old"))
  )
```

```{r}

drivers %>%
  filter(pan_cancer_driver == TRUE) %>%
  distinct(gene.refgene, Role) %>%
  count(gene.refgene) %>%
  filter(n > 1) 

```

```{r}

drivers %>%
  filter(pan_cancer_driver == TRUE) %>%
  distinct(gene.refgene, Role) %>%
  filter(gene.refgene %in% (
    drivers %>%
      filter(pan_cancer_driver == TRUE) %>%
      distinct(gene.refgene, Role) %>%
      count(gene.refgene) %>%
      filter(n > 1) %>%
      pull(gene.refgene)
  )) %>%
  arrange(gene.refgene) %>%
  head(20)

```


```{r}

drivers_vol <- drivers %>%
  filter(pan_cancer_driver == TRUE | is_driver_mut == TRUE | is_cosmic_gte_3 == TRUE) %>%
  select(gene.refgene, cancertype, oncogene, 
         tumor_suppressor, Role, pan_cancer_driver, sampleid, exonicfunc.refgene) %>%
  distinct(gene.refgene, sampleid, .keep_all = TRUE) # Explicit distinct() — removes rows where gene + sample are identical keeps the first occurrence of other columns like oncogene, Role etc. I don't want a hit at the same place twice.

# get gene names that appear in at least 10 samples pr cancertype
# store as a seperate
common_driver_genes <- drivers_vol %>%
  group_by(cancertype, gene.refgene) %>%
  summarise(n = n_distinct(sampleid)) %>%
  filter(n >= 10) %>%
  ungroup() %>%
  distinct(gene.refgene) %>%
  pull(gene.refgene)

# filter drivers_vol to only common genes
# now drivers_vol still has all its columns but only frequent genes
drivers_vol <- drivers_vol %>%
  filter(gene.refgene %in% common_driver_genes)

# some of them are both characterized as both oncogenes and TMR, because of this i assigned them as both.

# create a clean gene-level Role lookup
role_lookup <- drivers %>%
  filter(pan_cancer_driver == TRUE) %>%
  distinct(gene.refgene, Role) %>%
  group_by(gene.refgene) %>%
  summarise(Role_final = ifelse(n() > 1, "Both", as.character(first(Role))),
            .groups = "drop")

# join back to drivers_vol
drivers_vol <- drivers_vol %>%
  left_join(role_lookup, by = "gene.refgene") %>%
  select(-Role) 
```


```{r}

drivers_vol <- drivers_vol |> 
  mutate(exonicfunc.refgene = str_trim(exonicfunc.refgene))|>
  mutate(mutation_type = case_when(
  exonicfunc.refgene == "nonsynonymous SNV"        ~ "Missense",
  exonicfunc.refgene == "stopgain"                 ~ "Nonsense",
  exonicfunc.refgene == "stoploss"                 ~ "Nonstop",
  exonicfunc.refgene %in% c("frameshift deletion",
                             "frameshift insertion",
                             "frameshift substitution") ~ "Frameshift",
  exonicfunc.refgene %in% c("nonframeshift deletion",
                             "nonframeshift insertion",
                             "nonframeshift substitution") ~ "Inframe",
  is.na(exonicfunc.refgene)                        ~ NA_character_
))

```


### Oncomap

Making the mat. It needs a character matrix.

```{r}

# Values from `mutation_type` are not uniquely identified. I get this because toy cant have multiple mutations in the same gene. So pivot wider does not know which one to put in the cell. 
drivers_vol <- drivers_vol |>
  mutate(mutation_type = factor(mutation_type, 
levels = c("Nonsense", "Frameshift", "Inframe", "Missense", "Nonstop")))|> 
  group_by(gene.refgene, sampleid) |>
  slice_min(mutation_type) |>  # picks most severe (order of factors). The convention is to use the most severe one:"Nonsense > Frameshift > Splice > Missense > In-frame > other".
  ungroup()|>
  mutate(mutation_type = as.character(mutation_type))


# making the matrix for oncoprint.
mat <- drivers_vol |> 
  select(sampleid, gene.refgene, mutation_type) |> 
  pivot_wider(names_from = sampleid, id_cols = gene.refgene, values_from = mutation_type, values_fill = "", values_fn = first) |> 
  tibble::column_to_rownames("gene.refgene") |> 
  as.matrix()

mat[is.na(mat)] <- ""

# wanna find the top ones
 drivers_vol <- drivers_vol |> filter(!is.na(mutation_type)) # should i remove NA's?
 
top_genes <- drivers_vol |> 
  filter(cancertype == "BLCA") |> # trying with one cancertype first.
   group_by(gene.refgene) |>
   summarise(n = n_distinct(sampleid)) |>
   arrange(desc(n)) |>
   slice_head(n = 20) |>
   pull(gene.refgene)

# find BLCA sample IDs
blca_samples <- drivers_vol |>
  filter(cancertype == "BLCA") |>
  pull(sampleid) |>
  unique()

# Subset matrix til top genes OG kun BLCA samples
top_mat <- mat[top_genes, colnames(mat) %in% blca_samples]

# making a cancer type vector for column_split in oncoPrint. maybe i could use this one? 
cancertype_vec <- drivers_vol |> 
  select(sampleid, cancertype) |> 
  distinct() |>
  filter(sampleid %in% colnames(top_mat)) |>
  arrange(match(sampleid, colnames(top_mat))) |>
  pull(cancertype)


drivers_vol |> group_by(sampleid,cancertype) |> filter(cancertype == "BLCA")
clin |> group_by(sampleid,`cancer type abbreviation`) |> filter( `cancer type abbreviation` == "BLCA")

sum(drivers_vol$sampleid %in% clin$sampleid)

```

Joining clinical data. 

```{r}

clin_join <- clin |> 
  select(sampleid, bmi_groups, age_cat, gender, age_at_initial_pathologic_diagnosis) |>
  distinct(sampleid, .keep_all = TRUE)  # keep only first row per sampleid


drivers_vol <- drivers_vol |> 
  left_join(clin_join, by = 'sampleid', relationship = "many-to-many")


```


```{r}

# BMI
bmi_vec <- drivers_vol |>
  select(sampleid, bmi_groups) |>
  distinct() |>
  filter(sampleid %in% colnames(top_mat)) |>
  arrange(match(sampleid, colnames(top_mat))) |>
  pull(bmi_groups)

# Age
age_vec <- drivers_vol |>
  select(sampleid, age_cat) |>
  distinct() |>
  filter(sampleid %in% colnames(top_mat)) |>
  arrange(match(sampleid, colnames(top_mat))) |>
  pull(age_cat)

# Gender
gender_vec <- drivers_vol |>
  select(sampleid, gender) |>
  distinct() |>
  filter(sampleid %in% colnames(top_mat)) |>
  arrange(match(sampleid, colnames(top_mat))) |>
  pull(gender)


```


Making the alter_fun

```{r}
library(ComplexHeatmap)
suppressPackageStartupMessages(library(ComplexHeatmap))

col <- c(
  Missense   = "#008000",
  Nonsense   = "#000000", 
  Frameshift = "#FF0000",
  Inframe    = "#FFA500",
  Nonstop    = "#0000FF"
)

alter_fun = list(
  background = alter_graphic("rect", fill = "#CCCCCC"),
  Nonsense   = alter_graphic("rect", fill = col["Nonsense"]),
  Frameshift = alter_graphic("rect", fill = col["Frameshift"]),
  Inframe    = alter_graphic("rect", fill = col["Inframe"]),
  Missense   = alter_graphic("rect", fill = col["Missense"]),
  Nonstop    = alter_graphic("rect", fill = col["Nonstop"])
)

```


```{r}

column_title = "OncoPrint | TCGA | BLCA"
heatmap_legend_param = list(title = "Alternations", at = c("Missense","Nonsense", "Frameshift", "Inframe","Nonstop"), 
        labels = c("Missense","Nonsense", "Frameshift", "Inframe","Nonstop"))

oncoplot <- oncoPrint(top_mat,
          alter_fun = alter_fun,
          col = col,
          remove_empty_columns = TRUE,
          remove_empty_rows = TRUE) 


oncoplot2 <- oncoPrint(top_mat,
          alter_fun = alter_fun,
          col = col,
          column_split = bmi_vec,
          pct_side = "right", row_names_side = "left",
          top_annotation = HeatmapAnnotation(
            cbar = anno_oncoprint_barplot(),
            BMI    = bmi_vec,
            Age    = age_vec,
            Gender = gender_vec,
            col = list(
              BMI    = c(Normal = "green", Overweight = "orange", Obese = "red"),
              Age    = c(old = "lightblue", young = "darkblue"),
              Gender = c(MALE = "lightblue", FEMALE = "pink")
            )
          ),
          remove_empty_columns = TRUE)


pdf("figures/oncoplots/oncoprint_BLCA.pdf", width = 20, height = 10)
draw(oncoplot2)
dev.off()

```

```{r}
# removes NA
keep <- !is.na(bmi_vec) & !is.na(age_vec) & !is.na(gender_vec)

top_mat_clean <- top_mat[, keep]
bmi_vec_clean <- bmi_vec[keep]
age_vec_clean <- age_vec[keep]
gender_vec_clean <- gender_vec[keep]

oncoplot_bmi <- oncoPrint(top_mat_clean,
          alter_fun = alter_fun,
          col = col,
          column_split = bmi_vec_clean,
          pct_side = "right", row_names_side = "left",
          top_annotation = HeatmapAnnotation(
            cbar = anno_oncoprint_barplot(),
            BMI    = bmi_vec_clean,
            Age    = age_vec_clean,
            Gender = gender_vec_clean,
            col = list(
              BMI    = c(Normal = "green", Overweight = "orange", Obese = "red"),
              Age    = c(old = "lightblue", young = "darkblue"),
              Gender = c(MALE = "lightblue", FEMALE = "pink")
            )
          ),
          remove_empty_columns = TRUE)

oncoplot_age <- oncoPrint(top_mat_clean,
          alter_fun = alter_fun,
          col = col,
          column_split = age_vec_clean,
          pct_side = "right", row_names_side = "left",
          top_annotation = HeatmapAnnotation(
            cbar = anno_oncoprint_barplot(),
            BMI    = bmi_vec_clean,
            Age    = age_vec_clean,
            Gender = gender_vec_clean,
            col = list(
              BMI    = c(Normal = "green", Overweight = "orange", Obese = "red"),
              Age    = c(old = "lightblue", young = "darkblue"),
              Gender = c(MALE = "lightblue", FEMALE = "pink")
            )
          ),
          remove_empty_columns = TRUE)

pdf("figures/oncoplots/oncoprint_BLCA_bmi.pdf", width = 20, height = 10)
draw(oncoplot_bmi)
dev.off()

pdf("figures/oncoplots/oncoprint_BLCA_age.pdf", width = 20, height = 10)
draw(oncoplot_age)
dev.off()

```

```{r}
groups <- c("Normal", "Overweight", "Obese")

for (grp in groups) {
  keep <- !is.na(bmi_vec) & bmi_vec == grp
  
  op <- oncoPrint(top_mat[, keep],
            alter_fun = alter_fun,
            col = col,
            pct_side = "right", row_names_side = "left",
            column_title = grp,
            remove_empty_columns = TRUE,
            remove_empty_rows = TRUE)
  
  pdf(paste0("figures/oncoplots/oncoprint_BLCA_", grp, ".pdf"), width = 20, height = 10)
  draw(op)
  dev.off()
}

```

```{r}
# per cancer type
for (ct in unique(drivers_vol$cancertype)) {
  
  top_genes <- drivers_vol |>
    filter(cancertype == ct) |>
    group_by(gene.refgene) |>
    summarise(n = n_distinct(sampleid)) |>
    arrange(desc(n)) |>
    slice_head(n = 20) |>
    pull(gene.refgene)
  
  ct_samples <- drivers_vol |>
    filter(cancertype == ct) |>
    pull(sampleid) |>
    unique()
  
  if (length(top_genes) < 2 | length(ct_samples) < 2) next
  
  top_mat <- mat[top_genes[top_genes %in% rownames(mat)], 
                 colnames(mat) %in% ct_samples, drop = FALSE]
  
  bmi_vec    <- clin$bmi_groups[match(colnames(top_mat), clin$sampleid)]
  age_vec    <- clin$age_cat[match(colnames(top_mat), clin$sampleid)]
  gender_vec <- clin$gender[match(colnames(top_mat), clin$sampleid)]
  
  # BMI groups
  for (grp in c("Normal", "Overweight", "Obese")) {
    keep <- !is.na(bmi_vec) & bmi_vec == grp
    if (sum(keep) < 2) next
    op <- oncoPrint(top_mat[, keep, drop = FALSE],
              alter_fun = alter_fun, col = col,
              pct_side = "right", row_names_side = "left",
              column_title = grp,
              remove_empty_columns = TRUE,
              remove_empty_rows = TRUE)
    pdf(paste0("figures/oncoplots/", ct, "_BMI_", grp, ".pdf"), width = 20, height = 10)
    draw(op)
    dev.off()
  }
  
  # Age groups
  for (grp in c("young", "old")) {
    keep <- !is.na(age_vec) & age_vec == grp
    if (sum(keep) < 2) next
    op <- oncoPrint(top_mat[, keep, drop = FALSE],
              alter_fun = alter_fun, col = col,
              pct_side = "right", row_names_side = "left",
              column_title = grp,
              remove_empty_columns = TRUE,
              remove_empty_rows = TRUE)
    pdf(paste0("figures/oncoplots/", ct, "_Age_", grp, ".pdf"), width = 20, height = 10)
    draw(op)
    dev.off()
  }
  
  # Gender groups
  for (grp in c("MALE", "FEMALE")) {
    keep <- !is.na(gender_vec) & gender_vec == grp
    if (sum(keep) < 2) next
    op <- oncoPrint(top_mat[, keep, drop = FALSE],
              alter_fun = alter_fun, col = col,
              pct_side = "right", row_names_side = "left",
              column_title = grp,
              remove_empty_columns = TRUE,
              remove_empty_rows = TRUE)
    pdf(paste0("figures/oncoplots/", ct, "_Gender_", grp, ".pdf"), width = 20, height = 10)
    draw(op)
    dev.off()
  }
}

```



```{r}
for (ct in unique(drivers_vol$cancertype)) {
  ct_samples <- drivers_vol |>
    filter(cancertype == ct) |>
    pull(sampleid) |>
    unique()
  
  bmi_vec    <- clin$bmi_groups[match(ct_samples, clin$sampleid)]
  age_vec    <- clin$age_cat[match(ct_samples, clin$sampleid)]
  gender_vec <- clin$gender[match(ct_samples, clin$sampleid)]
  
  cat(ct, "- BMI:", table(bmi_vec), "| Age:", table(age_vec), "| Gender:", table(gender_vec), "\n")
}

```

